\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-14 paper.qmd
Update construct 2D model section

Hexagonal binning is a data visualization technique that aggregates high-dimensional data into a two-dimensional representation using hexagonal regions called bins [@article66]. The hexagonal shape is preferred due to its circular-like appearance, providing smoother transitions between neighboring bins and efficient data aggregation. This technique is especially useful for handling large data sets as it reduces visual clutter while capturing the underlying data distribution effectively. By considering the data distribution, important features are captured, and gaps between bins are minimized, leading to a more accurate representation of the data. In our algorithm, hexagonal binning is utilized to create a low-dimensional manifold.

@@

The first step involves to determine the optimal number of bins ($b$) for creating regular hexagons in the hexagonal grid. To achieve this, we rely on the relationship between the diameter ($h$) and the area ($A$) of regular hexagons, as given by @eq-equation3.

@@

To capture the data's structure effectively, we consider the optimal $A = 1$ (see @fig-binsize). Using @eq-equation4, we then calculate $h$ of a regular hexagon.

@@

  Next, we refer the `hexbin` function in R, which provides a relationship between the diameter of the hexagon ($h$) and the number of bins along the x-axis ($b_1$) (@eq-equation6).

@@

By utilizing the calculated $h$ and the range of the non-linear projection component 1 ($r_1$), we determine the $b_1$ (@eq-equation5). The result is rounded up to the nearest whole number to ensure an appropriate number of bins that capture the variability of the data along the x-axis.

@@

**(b) Determine the effective shape parameter** ($s$)

In this step, we determine the optimal shape parameter ($s$) for the hexagonal bins, which significantly influences their shape and arrangement within the grid. The shape parameter ($s$) in the hexagonal binning algorithm is defined as the ratio of the height ($y$) to the width ($x$) of the plotting region, referring to the `hexbin` function in R (see @eq-equation1). It determines the shape of the plotting regions and plays a vital role in generating an appropriate hexagonal grid for the data visualization.

@@

To calculate the effective shape parameter ($s$) for our algorithm, we consider the ranges of the non-linear projection components. Denoting the range of the non-linear projection component 1 as $r_1$ and the range of the non-linear projection component 2 as $r_2$, we find the shape parameter using @eq-equation2.

@@

@@

```{r}
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the same **s_curve_noise_umap** data: (a) b = (4, 8), s = 1.643542, (b) b = (12, 22), s = 1.643542, and (c) b = (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with yellow indicating higher point density and darker colors representing lower point density within each bin. Does  a value of number of bins exist to effectively represent the low-dimensional data?"

@@

  After computing hexagonal grid configurations, the 2D emdeddings are binned into hexagon cells. As a result of hexagonal binning , the bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2}$) (see @fig-meth Step 2) are obtained.

@@

  In this step, the algorithm proceeds to triangulate the centroids (see @fig-meth (Step 3)) of the hex bins. Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation. In our algorithm, triangulation helps us identify geometric relationships and patterns in the data.

@@

The total number of bins is a crucial parameter that defines the granularity of our hexagonal grid. This parameter particularly influential in capturing the intricacies of the data structure. It is determined by multiplying the number of bins along the x-axis (\(b_1\)) with the number of bins along the y-axis (\(b_2\)), as per the formula:

$$
b = b_1 \times b_2
$${#eq-equationp1}

  Here, \(b\) represents the total number of bins. Adjusting the parameter \(b_1\) provides control over the total bin count (\(b\)), allowing us to fine-tune the grid's configuration along the x-axis based on specific requirements.

>>>>

The hexagonal grid, formed through hexagonal binning [@Carr1987, @article66], serves as a type of bivariate histogram employed to visualize the structure of high-D data. Hexagons, being one of only three regular polygons capable of tessellating a plane [@Carr2013], possess both symmetry of nearest neighbors and the maximum number of sides for a regular tessellation of the plane. This unique combination makes hexagons more efficient in covering the plane compared to other regular tessellations. Additionally, hexagons exhibit lower visual bias when displaying densities, setting them apart from other regular tessellations [@Dan2023]. In our algorithm, hexagonal binning is used as the initial step of constructing the 2D model and the total number of bins ($b$) is the crucial parameter that defines the granularity of the hexagonal grid.

@@

First, the number of bins along the x-axis ($b_1$) is computed using the relationship between the diameter ($h$) and the area ($A$) of regular hexagons (see @eq-equation3).

@@

To construct regular hexagons, $A = 1$ (see @fig-binsize) use as the default setting. Then, the diameter ($h$) of the regular hexagons is calculated (see @eq-equation4).

@@

@Carr2013 mentioned about the relationship between the diameter ($h$) of regular hexagons and the height ($y$) of the plotting region. According to our algorithm, the height ($y$) of the plotting region is the the range of 2D embedding component 1 ($r_1$) (see @eq-equation6).

@@

After rearranging the @eq-equation6 as @eq-equation5, $b_1$ is computed. The $b_1$ value is rounded up to the nearest whole number to have an integer value.

@@

**(b) Determine the shape parameter** ($s$)

In this step, we determine the shape parameter ($s$) for the hexagonal bins, which significantly influences their shape and arrangement within the grid. The $s$ in the hexagonal binning algorithm is defined as the ratio of the height ($y$) to the width ($x$) of the plotting region as defined in @eq-equation1 [@Carr2013].

@@

The shape parameter ($s$) of our algorithm is calculated as the ratio of the ranges of 2D embedding components, where $r_1$ and $r_2$ represent the ranges of 2D embedding components 1 and component 2, respectively (see @eq-equation2).

@@

**(c) Determine the number of bins along the y-axis** ($b_2$)

Next, the number of bins along the y-axis is computed based on the number of bins along the x-axis ($b_1$) and the shape parameter ($s$) (see @eq-equation12) [@Carr2013].

$$
b_2 = 2 * \left(\frac{(b_1 \times s)}{\sqrt(3)} + 1.5001 \right)
$${#eq-equation12}

**(d) Determine the total number of bins** ($b$)

The total number of bins is determined by multiplying the number of bins along the x-axis ($b_1$) with the number of bins along the y-axis ($b_2$) (see @eq-equation13).

$$
b = b_1 \times b_2
$${#eq-equation13}

@@

```{r}
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. Does a value of number of bins exist to effectively represent the low-dimensional data?"

@@

As a result of hexagonal binning for high-D data, all the high-D data are clustered into hexagons. In this step, the bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2}$) (see @fig-meth Step 2) are obtained [@Carr2013].

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin centroids (see @fig-meth (Step 3)). Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26, @article54], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation.

@@

Adjusting the parameter $b_1$ provides control over the total number of bins $b$, allowing us to fine-tune the hexagonal grid.

### 2024-01-14 paper.tex
Update construct 2D model section

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

Hexagonal binning is a data visualization technique that aggregates
high-dimensional data into a two-dimensional representation using
hexagonal regions called bins \citep{article66}. The hexagonal shape is
preferred due to its circular-like appearance, providing smoother
transitions between neighboring bins and efficient data aggregation.
This technique is especially useful for handling large data sets as it
reduces visual clutter while capturing the underlying data distribution
effectively. By considering the data distribution, important features
are captured, and gaps between bins are minimized, leading to a more
accurate representation of the data. In our algorithm, hexagonal binning
is utilized to create a low-dimensional manifold.

@@

The first step involves to determine the optimal number of bins (\(b\))
for creating regular hexagons in the hexagonal grid. To achieve this, we
rely on the relationship between the diameter (\(h\)) and the area
(\(A\)) of regular hexagons, as given by Equation~\ref{eq-equation3}.

@@

To capture the data's structure effectively, we consider the optimal
\(A = 1\) (see Figure~\ref{fig-binsize}). Using
Equation~\ref{eq-equation4}, we then calculate \(h\) of a regular
hexagon.

@@

Next, we refer the \texttt{hexbin} function in R, which provides a
relationship between the diameter of the hexagon (\(h\)) and the number
of bins along the x-axis (\(b_1\)) (Equation~\ref{eq-equation6}).

@@

By utilizing the calculated \(h\) and the range of the non-linear
projection component 1 (\(r_1\)), we determine the \(b_1\)
(Equation~\ref{eq-equation5}). The result is rounded up to the nearest
whole number to ensure an appropriate number of bins that capture the
variability of the data along the x-axis.

@@

\textbf{(b) Determine the effective shape parameter} (\(s\))

In this step, we determine the optimal shape parameter (\(s\)) for the
hexagonal bins, which significantly influences their shape and
arrangement within the grid. The shape parameter (\(s\)) in the
hexagonal binning algorithm is defined as the ratio of the height
(\(y\)) to the width (\(x\)) of the plotting region, referring to the
\texttt{hexbin} function in R (see Equation~\ref{eq-equation1}). It
determines the shape of the plotting regions and plays a vital role in
generating an appropriate hexagonal grid for the data visualization.

@@

To calculate the effective shape parameter (\(s\)) for our algorithm, we
consider the ranges of the non-linear projection components. Denoting
the range of the non-linear projection component 1 as \(r_1\) and the
range of the non-linear projection component 2 as \(r_2\), we find the
shape parameter using Equation~\ref{eq-equation2}.

@@

@@

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the same \textbf{s\_curve\_noise\_umap} data: (a) b = (4, 8), s =
1.643542, (b) b = (12, 22), s = 1.643542, and (c) b = (21, 40), s =
1.643542. The hexbins are colored based on the density of points, with
yellow indicating higher point density and darker colors representing
lower point density within each bin. Does a value of number of bins
exist to effectively represent the low-dimensional data?}

@@

After computing hexagonal grid configurations, the 2D emdeddings are
binned into hexagon cells. As a result of hexagonal binning , the bin
centroids (\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2}\)) (see
Figure~\ref{fig-meth} Step 2) are obtained.

@@

In this step, the algorithm proceeds to triangulate the centroids (see
Figure~\ref{fig-meth} (Step 3)) of the hex bins. Triangulation is a
fundamental process in computational geometry and computer graphics that
involves dividing a set of points in a given space into interconnected
triangles \citep{article30}. One common algorithm used for triangulation
is Delaunay triangulation \citep{article26}, where points are connected
in a way that maximizes the minimum angles of the resulting triangles,
leading to a more regular and well-conditioned triangulation. In our
algorithm, triangulation helps us identify geometric relationships and
patterns in the data.

@@

The total number of bins is a crucial parameter that defines the
granularity of our hexagonal grid. This parameter particularly
influential in capturing the intricacies of the data structure. It is
determined by multiplying the number of bins along the x-axis ((b\_1))
with the number of bins along the y-axis ((b\_2)), as per the formula:

\begin{equation}\protect\hypertarget{eq-equationp1}{}{
b = b_1 \times b_2
}\label{eq-equationp1}\end{equation}

Here, (b) represents the total number of bins. Adjusting the parameter
(b\_1) provides control over the total bin count ((b)), allowing us to
fine-tune the grid's configuration along the x-axis based on specific
requirements.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

The hexagonal grid, formed through hexagonal binning
\citep[\citet{article66}]{Carr1987}, serves as a type of bivariate
histogram employed to visualize the structure of high-D data. Hexagons,
being one of only three regular polygons capable of tessellating a plane
\citep{Carr2013}, possess both symmetry of nearest neighbors and the
maximum number of sides for a regular tessellation of the plane. This
unique combination makes hexagons more efficient in covering the plane
compared to other regular tessellations. Additionally, hexagons exhibit
lower visual bias when displaying densities, setting them apart from
other regular tessellations \citep{Dan2023}. In our algorithm, hexagonal
binning is used as the initial step of constructing the 2D model and the
total number of bins (\(b\)) is the crucial parameter that defines the
granularity of the hexagonal grid.

@@

First, the number of bins along the x-axis (\(b_1\)) is computed using
the relationship between the diameter (\(h\)) and the area (\(A\)) of
regular hexagons (see Equation~\ref{eq-equation3}).

@@

To construct regular hexagons, \(A = 1\) (see Figure~\ref{fig-binsize})
use as the default setting. Then, the diameter (\(h\)) of the regular
hexagons is calculated (see Equation~\ref{eq-equation4}).

@@

\citet{Carr2013} mentioned about the relationship between the diameter
(\(h\)) of regular hexagons and the height (\(y\)) of the plotting
region. According to our algorithm, the height (\(y\)) of the plotting
region is the the range of 2D embedding component 1 (\(r_1\)) (see
Equation~\ref{eq-equation6}).

@@

After rearranging the Equation~\ref{eq-equation6} as
Equation~\ref{eq-equation5}, \(b_1\) is computed. The \(b_1\) value is
rounded up to the nearest whole number to have an integer value.

@@

\textbf{(b) Determine the shape parameter} (\(s\))

In this step, we determine the shape parameter (\(s\)) for the hexagonal
bins, which significantly influences their shape and arrangement within
the grid. The \(s\) in the hexagonal binning algorithm is defined as the
ratio of the height (\(y\)) to the width (\(x\)) of the plotting region
as defined in Equation~\ref{eq-equation1}.

@@

The shape parameter (\(s\)) of our algorithm is calculated as the ratio
of the ranges of 2D embedding components, where \(r_1\) and \(r_2\)
represent the ranges of 2D embedding components 1 and component 2,
respectively (see Equation~\ref{eq-equation2}).

@@

\textbf{(c) Determine the number of bins along the y-axis} (\(b_2\))

Next, the number of bins along the y-axis is computed based on the
number of bins along the x-axis (\(b_1\)) and the shape parameter
(\(s\)) (see Equation~\ref{eq-equation12}) \citep{Carr2013}.

\begin{equation}\protect\hypertarget{eq-equation12}{}{
b_2 = 2 * \left(\frac{(b_1 \times s)}{\sqrt(3)} + 1.5001 \right)
}\label{eq-equation12}\end{equation}

\textbf{(d) Determine the total number of bins} (\(b\))

The total number of bins is determined by multiplying the number of bins
along the x-axis (\(b_1\)) with the number of bins along the y-axis
(\(b_2\)) (see Equation~\ref{eq-equation13}).

\begin{equation}\protect\hypertarget{eq-equation13}{}{
b = b_1 \times b_2
}\label{eq-equation13}\end{equation}

@@

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the \textbf{UMAP} projections of \textbf{S-curve} data: (a) b = 32
(4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b =
840 (21, 40), s = 1.643542. The hexbins are colored based on the density
of points, with darker colors indicating higher point density and yellow
color representing lower point density within each bin. Does a value of
number of bins exist to effectively represent the low-dimensional data?}

@@

As a result of hexagonal binning for high-D data, all the high-D data
are clustered into hexagons. In this step, the bin centroids
(\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2}\)) (see Figure~\ref{fig-meth}
Step 2) are obtained \citep{Carr2013}.

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin
centroids (see Figure~\ref{fig-meth} (Step 3)). Triangulation is a
fundamental process in computational geometry and computer graphics that
involves dividing a set of points in a given space into interconnected
triangles \citep{article30}. One common algorithm used for triangulation
is Delaunay triangulation \citep[\citet{article54}]{article26}, where
points are connected in a way that maximizes the minimum angles of the
resulting triangles, leading to a more regular and well-conditioned
triangulation.

@@

Adjusting the parameter \(b_1\) provides control over the total number
of bins \(b\), allowing us to fine-tune the hexagonal grid.
